A pipeline to analyse time-course gene expression data

### 2019-11-19 scripts/manuscript.Rmd
Small fixes here and there

```{r echo=FALSE}
library(moanin)
```

@@

@@

```{r basis_matrix}
kable(head(moanin_model$basis))
```

@@

Another difficulty with such an approach is making sense of the general
temporal structure for any particular gene, as different genes will have
different combinations of timepoints DE. For the comparison of the Kawasaki
strain to the control, for example, there are `length(signifCombos)` genes
found DE in some timepoints, and there are `length(tabCombos)` different
combinations of timepoints for which they are DE. Some of these make sense,
such as DE in timepoints 48H-168H (`tabCombos["48,60,72,120,168"]` genes), but
many are very fragmentary. For example there are `tabCombos["48,60,120,168"]`
genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H. Many of
these genes are likely to have not made the cutoff for significance in 72H,
but don't show real differences in the overall trend between 48H-168H.

@@

```{r showMissingDEGenes, echo=FALSE}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
plot_splines_data(data, moanin_model, subset_data=exampleGenes,
		  colors=ann_colors$Group)
```

@@

| Name   | Formula   |     |
|------- | --------- |--- |
| timely | $\text{lfc}(t)$  | Function of time |
| sum  | $\sum_t lfc(t)$  | Sum of log fold change. |
| abs_sum  | $\sum_t \lvert lfc(t)\lvert$  | Always positive |
| max  | $\max_t \lvert lfc(t) \lvert$  | Always positive  |
| min  | $\min_t \lvert lfc(t) \lvert$  | Always positive |
| timecourse | See details below | Captures overall strength and direction  |

@@

$$\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T lfc(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {lfc(t)} \lvert \right)$$

@@

@@

plot(lfc_timecourse, -log10(pvalue), pch=20, main="Volcano plot",
     xlim=c(-2.5, 2))
```

@@

@@

```{r assigning_scores_and_labels}
# Then assign scores and labels to all the data, using a goodness-of-fit
# scoring function.
scores_and_labels = splines_kmeans_score_and_label(
    data, kmeans_clusters)
labels = scores_and_labels$labels

# Let's keep only the list of genes that have a label.
scores = rowMin(scores_and_labels$scores[!is.na(labels), ])
labels = unlist(labels[!is.na(labels)])
genes_in_cluster2 = names(labels[labels==cluster_of_interest])
```

Let us now, for each cluster, visualize the distribution of scores of the
genes assigned to that cluster.

```{r visualize_scores, fig.height=6, fix.width=6}
# Get the best score for each gene
par(mfrow=c(4, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(scores[labels == cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
}

@@

@@

**EAP: I don't know what to think about these plots -- these are best scoring,
but pretty noisy. Also the per-gene spline is sometimes not as good of a
explanation as the cluster centroid. It's seems like an important plot to
make**-

**NV: this were the worst scoring genes. I fixed this and the plots now look
much better**

@@

**EAP: `plot_model_explorer` is relatively slow, because it does all those pairwise calculations. And then to do it for smaller subset, redo all the calculations. Can't the output be saved, or the calculations be a different step from the plotting?**

>>>>

@@

## Package versions

The following packages are needed for this workflow.

```{r all_libraries}
# From CRAN
library(NMF)
library(ggfortify)

# From Bioconductor
library(topGO)
library(biomaRt)

# From GitHub
library(moanin)
```

To install `moanin` directly from GitHub, use the devtools' `install_github`
function:

@@

```{r basis_matrix}
kable(head(moanin_model$basis), digits=2) %>%
  kable_styling(font_size = 7)
```

@@

Another difficulty with such an approach is making sense of the general
temporal structure for any particular gene, as different genes will have
different combinations of timepoints DE. For the comparison of the Kawasaki
strain to the control, for example, there are `r length(signifCombos)` genes
found DE in some timepoints, and there are `r length(tabCombos)` different
combinations of timepoints for which they are DE. Some of these make sense,
such as DE in timepoints 48H-168H (`r tabCombos["48,60,72,120,168"]` genes), but
many are very fragmentary. For example there are `r tabCombos["48,60,120,168"]`
genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H. Many of
these genes are likely to have not made the cutoff for significance in 72H,
but don't show real differences in the overall trend between 48H-168H.

@@

```{r showMissingDEGenes, echo=FALSE}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
plot_splines_data(data, moanin_model, subset_data=exampleGenes,
		  colors=ann_colors$Group, smooth=TRUE)
```

@@

| Name   | Formula   |     |
|------- | --------- |--- |
| timely | $\text{lfc}(t)$  | Function of time |
| sum  | $\sum_t \text{lfc}(t)$  | Sum of log fold change. |
| abs_sum  | $\sum_t \lvert \text{lfc}(t)\lvert$  | Always positive |
| max  | $\max_t \lvert \text{lfc}(t) \lvert$  | Always positive  |
| min  | $\min_t \lvert \text{lfc}(t) \lvert$  | Always positive |
| timecourse | See details below | Captures overall strength and direction  |

@@

$$\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T \text{lfc}(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {\text{lfc}(t)} \lvert \right)$$

@@

The returning object is a matrix, where each row corresponds to a gene,
each column to a contrasts, and each entry to the log-fold change for this
pair of contrast and gene.

```{r display_lfc}
kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE, escape=F)
```

Each methods of computing the log-fold change captures different elements of
the time-course data: overall change, largest change, ...

@@

plot(lfc_timecourse, -log10(pvalue), pch=20, main="Volcano plot",
     xlim=c(-2.5, 2), xlab="Timecourse lfc")
```

@@

The `splines_kmeans` function returns a named list with:

- `centroids`: a matrix containing the cluster centroids. The matrix is of
  shape (`n_centroid`, `n_samples`).
- `clusters`: a vector of size `n_genes`, containing the

@@

```{r assigning_scores_and_labels}
# Then assign scores and labels to all the data, using a goodness-of-fit
# scoring function.
scores_and_labels = splines_kmeans_score_and_label(
    data, kmeans_clusters)
```

The `scores_and_labels` list contains two elements:

    - `scores_and_labels$scores`: the matrix of shape `n_cluster` x `n_genes`,
      containing for each gene and each cluster, the goodness of fit as
      described above.
    - `scores_and_labels$labels`: the labels for all of the genes with a
      good-enough goodness-of-fit score.

First, let us visualize the distribution of goodness-of-fit scores for each
cluster. Here, we display the whole set of scores assigned to each cluster,
prior to filtering poor-quality fits.

```{r visualize_scores, fig.height=6, fix.width=6, echo=TRUE, caption="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
# Get the best score and best label for all of the genes
scores = rowMin(scores_and_labels$scores)
labels = apply(scores_and_labels$scores, 1, which.min)
max_score = quantile(scores, 0.5)
par(mfrow=c(3, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(scores[labels == cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
    abline(v=max_score, col="red", lwd=3, lty=2)
}

@@

Note that for some of the genes, the best scores is 1. A score of 1 indicates
that there is no correlation between the gene and the cluster: those genes fit
poorly to all clusters and the assignment to one cluster is done randomly..
This underlines the importance of filtering genes that fit poorly to clusters.

```{r scores_and_labels}
labels = scores_and_labels$labels

# Let's keep only the list of genes that have a label.
labels = unlist(labels[!is.na(labels)])

# And also keep track of all the genes in cluster 2.
genes_in_cluster2 = names(labels[labels==cluster_of_interest])
```

@@

@@
